---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- say directory, not folder -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# notestar 📓⭐

<!-- badges: start -->
<!-- badges: end -->

notestar is a notebook system built on the targets package: *notes* with
*tar*gets. I originally called this package tarnotes, but that sounded gross.

**This README and the package are under development.**

## Installation

You can install notestar from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tjmahr/notestar")
```

## A demo notebook

Here is an example project/notebook showing how notestar combines various .Rmd
into a single HTML file: <https://github.com/tjmahr/notestar-demo>.

## A small example

notestar works best inside of a data analysis project and specifically, as a
part of an RStudio project. That is, we have some directory for our project.
Everything we do or create will live in that directory, and that directory is
the default working directory for all of our R code.

Let's create a new directory inside of a temporary directory and make that the
home base for our project.

```{r}
project_dir <- file.path(tempdir(), pattern = "my-project")
dir.create(project_dir)
setwd(project_dir)
```

```{r, echo = FALSE}
knitr::opts_knit$set(root.dir = project_dir)
```


Nothing here!

```{r}
fs::dir_tree(all = TRUE)
```

`use_notestar()` will populate the project folder with some boilerplate files.

```{r}
library(notestar)
use_notestar()
```

Now, we have the basic skeleton. 

```{r}
fs::dir_tree(all = TRUE)
```

These will be documented in detail below, but we
have a folder `notebook` where we store our notebook entries (as RMarkdown
files), `notebook/book/` where we store the knitted versions of those entries
(as markdown files), and `_targets` which orchestrates the compilation of the
notebook.

`targets::tar_make()` will then compile the notebook by:

  - knitting each Rmd file in `notebook` *if* necessary
  - collating the md files in `notebook/book/` into a single-document bookdown
    book with bookdown/RMarkdown/pandoc (*if* necessary).

I say "*if* necessary" because targets only builds targets if the target has not
been built yet or if the target is out of data. Thus, notestar doesn't waste
time regenerating earlier entries if they have not changed.


Here we build the notebook and see targets build each target.

```{r}
targets::tar_make()
```

If we ask it to build the book again, it skips everything but a special
spellchecking target set to always run.

```{r}
targets::tar_make()
```

Right now, our compiled notebook (`"notebook/book/docs/notebook.html"`) is just
the title page:

```{r shot1, out.width = "35%", echo = FALSE}
webshot::webshot(
  "notebook/book/docs/notebook.html",  
  file = "shot1.png",
  vwidth = 400, 
  vheight = 400,
  zoom = 2
)
```

If we look at the project tree, we see some additions.

```{r}
fs::dir_tree(all = TRUE)
```

Briefly, there are some md files in `notebook/book/` as well as some
bookdown-related files (.yaml files, .rds file). There is also the output of
bookdown in `notebook/book/docs`. `_targets/` is a new directory. It is the
object and metadata storage for targets.



We can create a new entry from a template using `notebook_create_page()` and
regenerate the notebook. (A slug is some words we include in the filename to
help remember what the entry is about.)

```{r}
notebook_create_page(slug = "hello-world")
```

Now targets has to rebuild the notebook because there is a new entry that needs
to be folded in.

```{r}
targets::tar_make()
```

That entry appears in the notebook.

```{r shot2, out.width = "35%", echo = FALSE}
webshot::webshot(
  "notebook/book/docs/notebook.html",
  file = "shot2.png",
  vwidth = 400, 
  vheight = 500, 
  zoom = 2
)
```

From here, we go with the flow. We use targets as we normally would, modifying
`R/functions.R` and `targets.R` to set up our data-processing pipeline. We can
now use our notebook to do reporting and exploration as part of our
data-processing pipeline. Things we make with targets can be `tar_read()` into
our notebook entries and tracked as dependencies.

```{r, echo = FALSE}
knitr::opts_knit$set(root.dir = NULL)
```




## How it all works [todo]

targets + bookdown + some tricks

```{r, eval = FALSE}
fs::dir_tree(path = project_dir, all = TRUE)
```

  - `.here` is a sentinel file for the here package. It indicates where the
    project root is located.
  - `notebook/` houses the .Rmd files that become entries in the notebook. By
    default, it includes `index.Rmd` (a bookdown title page) and
    `0000-00-00-references.Rmd` (an entry for a pandoc bibliography provided at
    the end/bottom of the notebook).
  - `notebook/book/` houses the knitted versions of the `.Rmd` entries and a
    timestamped version of the `index.Rmd`. There is a special file here
    `knitr-helpers.R` which is run before the Rmd -\> md and before the md -\>
    notebook stage. `assets/` is where knitted images go and where other assets
    to include go.
  - `_targets.R` stores the workflow for building the notebook.
  - `R/functions.R` creates

