---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- say directory, not folder -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# notestar 📓⭐

<!-- badges: start -->
[![R-CMD-check](https://github.com/tjmahr/notestar/workflows/R-CMD-check/badge.svg)](https://github.com/tjmahr/notestar/actions)
<!-- badges: end -->

notestar is a notebook system built on the targets package: *notes* with
*tar*gets.

**This README and the package are under development.**

## Installation

You can install notestar from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tjmahr/notestar")
```

## A demo notebook

Here is an example project/notebook showing how notestar combines
various .Rmd files into a single HTML file:
<https://github.com/tjmahr/notestar-demo>.

## A small example

notestar works best inside of a data analysis project and specifically, as a
part of an RStudio project. That is, we have some directory for our project.
Everything we do or create will live in that directory, and that directory is
the default working directory for all of our R code.

For demonstration, let's create a new directory inside of a temporary
directory and make that the home base for our project.

```{r}
project_dir <- file.path(tempdir(), pattern = "my-project")
dir.create(project_dir)
setwd(project_dir)
```

```{r, echo = FALSE}
knitr::opts_knit$set(root.dir = project_dir)
```

Nothing here!

```{r}
fs::dir_tree(all = TRUE)
```

### Initial file skeleton

`use_notestar()` will populate the project folder with some boilerplate files.

```{r}
library(notestar)
use_notestar()
```

Now, we have the basic skeleton. 

```{r}
fs::dir_tree(all = TRUE)
```

The file **`config.yml`** is a config-package configuration file. It
specifies the main folder and file locations for our notebook. Each of
these is described by a `comment` field.

```{r}
readLines("config.yml") |>
  writeLines()
```

Two .Rmd files are automatically included: `index.Rmd` and
`0000-00-00.Rmd`. These are the first and last entries (top and bottom
parts) of the notebook. **`index.Rmd`** houses metadata for the
notebook:

```{r}
readLines("notebook/index.Rmd") |>
  writeLines()
```

The yaml metadata in `index.Rmd` is created automatically inside the
`_targets.R` file. More on that later.

**`0000-00-00.Rmd`** is not meant to be edited. As it tells us, it
provides a "References" heading. When the bibliography is appended to
the end of the notebook, it will be printed under this heading.

```{r}
readLines("notebook/0000-00-00-references.Rmd") |>
  writeLines()
```

The file **`_targets.R`** orchestrates the compilation of the notebook using
the targets package. `targets::tar_make()` compiles the notebook by:

  - knitting each .Rmd file in `notebook` *if* necessary to produce a
    corresponding .md file `notebook/book/`.
  - collating the .md files in `notebook/book/` into a single-document
    bookdown book with bookdown/RMarkdown/pandoc (*if* necessary).

I say "*if* necessary" because targets only builds the targets in
workflow if the target has not been built yet or if the target is out of
date. Thus, notestar doesn't waste time regenerating earlier entries if
they or their dependencies have not changed.

Finally, **`.here`** is a sentinel file for the here package. It indicates where the
project root is located. **`R/functions.R`** is an (as-yet empty) R script that is `source()`-ed
at the start of `_targets.R`.

### Building, rebuilding

Here we build the notebook and see targets build each target.

```{r}
targets::tar_make()
```

If we ask it to build the book again, it skips everything---none of the
dependencies have changed---but a special spellchecking target set to
always run.

```{r}
targets::tar_make()
```

Right now, our compiled notebook (`"notebook/book/docs/notebook.html"`)
is just the title page:

```{r shot1, out.width = "35%", echo = FALSE}
webshot::webshot(
  "notebook/book/docs/notebook.html",  
  file = "shot1.png",
  vwidth = 400, 
  vheight = 400,
  zoom = 2
)
```

If we look at the project tree, we see some additions.

```{r}
fs::dir_tree(all = TRUE)
```

`_targets/` is a new directory. It is the object and metadata storage
for targets. We don't worry about it.

There are some md files in `notebook/book/` as well as some
bookdown-related files (`_bookdown.yml`, `_output.yml` and
`notebook.rds` file). There is also the output of bookdown in
`notebook/book/docs`. (`notebook/book/docs/notebook.html` is the file we
screenshotted earlier.)

We can create a new entry from a template using `notebook_create_page()` and
regenerate the notebook. (A slug is some words we include in the filename to
help remember what the entry is about.)

```{r}
notebook_create_page(date = "2022-02-22", slug = "hello-world")
```


Now targets has to rebuild the notebook because there is a new entry
that needs to be folded in. The network diagram shows that
`entry_2022_02_hello_world_rmd` is outdated (blue) so everything
downstream from it is also outdated.

```{r network1, dpi = 300}
targets::tar_visnetwork(targets_only = TRUE)
```


When we rebuild the notebook, that entry now appears in 
the HTML file.

```{r shot2, out.width = "35%", echo = 1:2, results = "hide"}
targets::tar_make()
#> [output omitted]

webshot::webshot(
  "notebook/book/docs/notebook.html",
  file = "shot2.png",
  vwidth = 400, 
  vheight = 500, 
  zoom = 2
)
```

#### Next steps

From here, we go with the flow. We use targets as we normally would, modifying
`R/functions.R` and `targets.R` to set up our data-processing pipeline. We can
now use our notebook to do reporting and exploration as part of our
data-processing pipeline. Things we make with targets can be `tar_read()` into
our notebook entries and tracked as dependencies.



### Finer point 1: .Rmd/.md files have hidden timestamps

Here is what a minimal Rmd file entry looks like:

```{r}
writeLines(readLines("notebook/2022-02-22-hello-world.Rmd"))
```

That first `<!--- comment --->` line on top is an HTML comment. It will
not be displayed when we view the final html file, but when the .Rmd file is knitted to produce the
corresponding .md, the timestamp will be updated. Here is the first line
of that .md file:

```{r}
writeLines(readLines("notebook/book/2022-02-22-hello-world.md")[1])
```

This timestamp allows us to mark a notebook entry as outdated even if
none of the text in the .md file has changed. Here is a motivating
example. Let's append a code chunk to the bottom of the notebook entry.
It will plot a histogram.

```{r}
entry_v0 <- readLines("notebook/2022-02-22-hello-world.Rmd")[1:13]
writeLines(
  c(
    entry_v0,
    "Would you look at all these 4's?",
    "```{r old-faithful, fig.width = 4, fig.height = 4}",
    "hist(faithful$eruptions)",
    "```"
  ),
  "notebook/2022-02-22-hello-world.Rmd"
)
```

And then we regenerate the notebook.


```{r shot3, out.width = "35%", echo = 1:2, results = "hide"}
targets::tar_make()
#> [output omitted]

webshot::webshot(
  "notebook/book/docs/notebook.html",  
  file = "shot3.png",
  vwidth = 400, 
  vheight = 400,
  zoom = 2
)
```

Let's store the current .md file lines so we can compare it to a later
version.

```{r}
entry_v1 <- readLines("notebook/book/2022-02-22-hello-world.md")
```

Now, suppose we wanted to change size or resolution of the plot. In this case, 
we will change the `fig.width` and `fig.height` values to 6 here.

```{r}
writeLines(
  c(
    entry_v0,
    "Would you look at all these 4's?",
    "```{r old-faithful, fig.width = 6, fig.height = 6}",
    "hist(faithful$eruptions)",
    "```"
  ),
  "notebook/2022-02-22-hello-world.Rmd"
)
```

And then we regenerate the notebook.

```{r shot4, out.width = "35%", echo = 1:2, results = "hide"}
targets::tar_make()
#> [output omitted]

webshot::webshot(
  "notebook/book/docs/notebook.html",  
  file = "shot4.png",
  vwidth = 400, 
  vheight = 400,
  zoom = 2
)
```


So now the figures image files have definitely changed. They are
different sizes. The text in the plots in the two screenshots are
different sizes. But the text of the .md files is the same---except for
the timestamp.

```{r}
entry_v2 <- readLines("notebook/book/2022-02-22-hello-world.md")
entry_v1 == entry_v2
entry_v1[1] 
entry_v2[1]
```

This phenomenon, where a change to an .Rmd file would not cause a change
in the text of a .md file, is the reason for the timestamp at the top of
the .Rmd file.


### Finer point 2: Unused figures are automatically deleted

What happens when you remove a figure from .Rmd file? Normally, it
sticks around, and you eventually find yourself with all kinds of old,
no-longer used figures. When knitting a .Rmd file, notestar removes all
existing figures associated with an entry beforehand. As a result, only 
figures that were created during the most recent knitting are retained.

Let's demonstrate this feature. Here are the current notebook assets:

```{r}
fs::dir_tree("./notebook/book/assets")
```

We will restore the original version of the entry so that the plot is no
longer created.

```{r, results = "hide"}
writeLines(entry_v0, "notebook/2022-02-22-hello-world.Rmd")

targets::tar_make()
#> [output omitted]
```

What we have now is an empty folder.

```{r}
fs::dir_tree("./notebook/book/assets")
```





```{r, echo = FALSE}
knitr::opts_knit$set(root.dir = NULL)
```

## How it all works [todo]

targets + bookdown + some tricks
