---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# notestar 📓⭐

<!-- badges: start -->
<!-- badges: end -->

notestar is a notebook system built on the targets package: *notes* with
*tar*gets. I originally called this package tarnotes, but that sounded gross.

**This README and the package are under development.**

## Installation

You can install notestar from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tjmahr/notestar")
```

## A small example

notestar works best inside of a data analysis project and specifically, as a
part of an RStudio project. That is, we have some folder for our project,
everything we do or create will live in the somewhere folder, and that folder is
the default working folder for all of our R code.

Let's create a new folder inside of a temporary directory and make that our
project.

```{r}
project_dir <- file.path(tempdir(), pattern = "my-project")
dir.create(project_dir)
setwd(project_dir)
```

```{r, echo = FALSE}
knitr::opts_knit$set(root.dir = project_dir)
```


`use_notestar()` will populate the project folder with some boilerplate files.

```{r}
library(notestar)
use_notestar()
```

`targets::tar_make()` will then create the notebook. 

```{r}
targets::tar_make()
```

Right now, it's just the title page:

```{r shot1, out.width = "35%", echo = FALSE}
webshot::webshot(
  "notebook/book/docs/notebook.html",  
  file = "shot1.png",
  vwidth = 400, 
  vheight = 400,
  zoom = 2
)
```

We can create a new entry from a template using `notebook_create_page()` and
regenerate the notebook. (A slug is some words we include in the filename to
help remember what the entry is about.)

```{r}
notebook_create_page(slug = "hello-world")
targets::tar_make()
```

Now, that entry appears in the notebook.

```{r shot2, out.width = "35%", echo = FALSE}
webshot::webshot(
  "notebook/book/docs/notebook.html",
  file = "shot2.png",
  vwidth = 400, 
  vheight = 500, 
  zoom = 2
)
```

```{r, echo = FALSE}
knitr::opts_knit$set(root.dir = NULL)
```

## How it all works

targets + bookdown + some tricks

## scratch paper

```{r, eval = FALSE}
fs::dir_tree(path = project_dir, all = TRUE)
```

  - `.here` is a sentinel file for the here package. It indicates where the
    project root is located.
  - `notebook/` houses the .Rmd files that become entries in the notebook. By
    default, it includes `index.Rmd` (a bookdown title page) and
    `0000-00-00-references.Rmd` (an entry for a pandoc bibliography provided at
    the end/bottom of the notebook).
  - `notebook/book/` houses the knitted versions of the `.Rmd` entries and a
    timestamped version of the `index.Rmd`. There is a special file here
    `knitr-helpers.R` which is run before the Rmd -\> md and before the md -\>
    notebook stage. `assets/` is where knitted images go and where other assets
    to include go.
  - `_targets.R` stores the workflow for building the notebook.
  - `R/functions.R` creates

